Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-16 submission/manuscript.Rmd
res 1 edits

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We systematically tested whether other tissue-specific genes are misexpressed in the brain with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].

@@

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO (5CKO) brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

>>>>

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus systematically tested whether other tissue-specific genes are misexpressed in the male brain with constitutive knockout of _Kdm5c_ (_Kdm5c^-/y^_, 5CKO)[@vallianatosMutuallySuppressiveRoles2020] by using a published list of mouse tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017].

@@

We found a large proportion of significantly upregulated genes (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1) within the male _Kdm5c_-KO amygdala and hippocampus are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched differentially expressed genes (DEGs) were testis genes (Figure 1A-B). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). An example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

@@

Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and muscles (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).
